Towards Correct Legion

### 2017-08-17 correctnesshpc/paper.tex
Added a little more background

Legion is a programming model for high performance computing that enables
reasoning about locality and concurrency via logical regions. It has the
attractive property that it is guaranteed to preserve serial semantics for
memory operations within logical regions. This allows programmers to avoid the
burden of reasoning about the non-determinism that arises from concurrent
computation. For this paper, we'll be formalizating work on from Triecher et.
al \cite{ooplsa13} While there has been much work extending Legion since then,
the core semantics has stayed relatively stable \cite{regent, deppart}.

Improvements to proof assistants over recent decades have made them a powerful
tool for implementing and reasoning about code. From provably correct compilers
to provably correct operating systems, proof assistants enable extremely high
certainty in correctness even for large scale software projects.

Coq is a dependently typed proof assistant. At its core is the Calculus of
Inductive Constructions, a small dependently typed programming language and
logic, capable of powerful abstractions for formal proofs. In the last 10 years 
great strides have been made in using it to both build systems and prove useful
properties about them \cite{compcert, certicoq, chlipala...}.

@@

One thing that makes Legion unique among high performance runtime systems is
that it provides a specification of its operational semantics \cite{oopsla13}.
We take this specification of and formalize it in Coq, enabling us to prove
lemmas and theorems about it. See Figure~\ref{?} for the Coq implementation of
the syntax, and Figure~\ref{semantics}and semantics. All code is available
online at \texttt{https://github.com/lanl/formallegion}.

@@

The Legion semantics are defined as a small step operational semantics, which
result in non-deterministic orderings of memory reads and writes. By defining a
notion of \texttt{valid\_interleaves}, we can reason formally about what evaluations 
can happen concurrently. This should allow us to prove formally that if the
runtime makes scheduling decisions based on region permissions in a particular
way, serial semantics are preserved with respect to the resulting heap state
and return value. This is a desirable property: it means that any code using
Legion can reason, either formally or informally, about its code with respect
to a simple serial semantics, and have those semantics be preserved for
parallel execution. This is in contrast to programming models like Cilk, which
claim a weaker property: a serial execution is one of many valid executions.

@@

One important property when proving correctness in Legion is to show that the
relations are equivalent to their closure, with respect to what they imply
about locations in those regions.  Figure~\ref{regionrelproof} for the
statement of this proof. Intuitively, this means that if a region is a
subregion of another region, and that region is a subregion of a third, then
the first subregion is a subregion of the third. Disjointness is commutative, 
and subregions of disjoint regions are disjoint.

>>>>

Legion is a programming model for high performance computing that enables
reasoning about locality and concurrency via logical regions. It has the
attractive property that it is guaranteed to preserve serial semantics for
memory operations within logical regions. This allows programmers to avoid the
burden of reasoning about the non-determinism that arises from concurrent
computation. For this paper, we'll be formalizating work on from Triecher et.
al \cite{ooplsa13}. While there has been much work extending Legion since then,
the core semantics has stayed relatively stable \cite{regent, deppart}.

By partitioning regions into \emph{subregions}, and tracking \emph{privileges}, 
Legion is able to ensure that certain code won't interfere or alias, and can
therefor be safely, at least with respect to those memory regions, run in
parallel. This theory has been put into practice to impressive effect, and has
been shown to scale both within a node and between \cite{legionpapers}. That
said, there are still challenges subtle bugs that occasionally show up in the
implementation. It is possible that these kinds of bugs could manifest
themselves in ways that would lead to incorrect conclusions when informing
major decisions, something we'd generally like to avoid. One goal of this paper
is to move towards formally and systematically removing \emph{all} bugs from
the programming model, giving everyone who uses is far more certainty that 
their code is correct.

To move towards this goal, we turn to mechanized proof assistants.
Improvements to proof assistants over recent decades have made them a powerful
tool for implementing and reasoning about code. From provably correct compilers
to provably correct operating systems, proof assistants enable extremely high
certainty in correctness even for large scale software projects.

Coq, in particular, has proven itself to be a particularly potent dependently
typed proof assistant. At its core is the Calculus of Inductive Constructions,
a small dependently typed programming language and logic, capable of powerful
abstractions for formal proofs \cite{cic}. In the last 10 years great strides
have been made in using it to both build systems and prove useful properties
about them \cite{compcert, certicoq, chlipala...}.

@@

One thing that makes Legion unique among high performance runtime systems is
that it provides a specification of its operational semantics \cite{oopsla13}.
We take this specification of and formalize it in Coq, enabling us to prove
lemmas and theorems about it. See Figure~\ref{syntax} for the Coq
implementation of the syntax, and Figure~\ref{semantics}and semantics. All code
is available online at \texttt{https://github.com/lanl/formallegion}.

@@

The Legion semantics are defined as a big step operational semantics, which
result in non-deterministic orderings of memory reads and writes. By defining a
notion of \texttt{valid\_interleaves}, we can reason formally about what evaluations 
can happen concurrently. This should allow us to prove formally that if the
runtime makes scheduling decisions based on region permissions in a particular
way, serial semantics are preserved with respect to the resulting heap state
and return value. This is a desirable property: it means that any code using
Legion can reason, either formally or informally, about its code with respect
to a simple serial semantics, and have those semantics be preserved for
parallel execution. This is in contrast to programming models like Cilk, which
claim a weaker property: a serial execution is one of many valid executions.

@@

One important property when proving correctness in Legion is to show that the
relations are equivalent to their closure, with respect to what they imply
about locations in those regions.  Figure~\ref{regionrelproof} for the
statement of this proof. Intuitively, this means that if a region is a
subregion of another region, and that region is a subregion of a third, then
the first subregion is a subregion of the third. Disjointness is commutative, 
and subregions of disjoint regions are disjoint. We've left out the Coq tactic
proofs for space, as they are likely less informative to the reader than the
propositions and definitions, but they can be found online as linked above. 
Lemma \texttt{resp\_clos\_sub} proves that given that if a subregion relation
holds with respect to a mapping \texttt{M} and a set of region constraints
\textOmega{Ω} for all regions \texttt{r1, r2} then holds for the closure of that
set of region constraints under that mapping. Given \texttt{resp\_clos\_sub}, 
Lemma \texttt{resp\_clos} proves that this implies that a mapping respects
a set of region constraints if and onlyo if it respects the closure of that
set. This non-trivial property is crucial for proving that the final decisions 
made by the tasking system based on the sets of constraints preserves 
serial semantics.
